---
title: "XCT.read: reading tree-ring width and density data"
subtitle: "A practical guide to the XCT.read R function"
author:
  - name: Louis Verschuren
  - name: Vladimir Matskovsky
  - name: Jan Van den Bulcke
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    smooth-scroll: true
    fig-responsive: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
editor: visual
---

## What does `XCT.read()` do?

`XCT.read()` reads the text files exported by **RingIndicator** and converts them into analysis-ready tree-ring width and X-ray density data.

The function can return:

- annual **ring width**;
- an annual **density parameter** calculated from a selected part of each ring;
- **ring width and density together** in long format;
- the complete **within-ring density profile**; or
- **MXD** values exported from RingIndicator area-of-interest analyses.

The main advantage is that the same RingIndicator exports can be used for several density definitions without having to re-export or manually manipulate the profiles.

::: callout-tip
## Most users can start here

If you want ring width plus the mean density of the final quarter of every ring, the default calculation is almost enough:

``` r
Data <- XCT.read(
  path = "Datafolder",
  output = "ringwidth_density",
  densityType = "fraction",
  area = c(0.75, 1),
  fun = "mean"
)
```
:::

## Where does it fit in the XCT workflow? hide the code of mermaid

```{mermaid}
%%| echo: false
flowchart LR
    A[X-ray micro-CT data] --> B[CoreProcessor]
    B --> C[RingIndicator]
    C --> D[Ring-width indications]
    C --> E[Corrected density profiles]
    C --> F[Corrected ring positions]
    D --> G[XCT.read]
    E --> G
    F --> G
    G --> H[Ring width]
    G --> I[Annual density]
    G --> J[Ring width + density]
    G --> K[Within-ring density profiles]
    C --> L[Area-of-interest MXD files]
    L --> G
```

`XCT.read()` expects files belonging to the same samples to be together in one folder. For the standard density calculations, filenames are matched by their sample prefix, for example:

``` text
L-722-1_ringwidth.txt
L-722-1_density_corr.txt
L-722-1_zpos_corr.txt
```

The included `Datafolder` contains example exports that can be used to run this document.

## Setup

Install the required R packages if needed:

``` r
install.packages(c("tidyverse", "dplR", "knitr"))
```

Load the packages and source the function:

```{r}
#| label: setup
library(tidyverse)
library(dplR)
library(knitr)

source("XCT.Read.R")

data_path <- "Datafolder"
```

::: callout-note
The examples below assume that `XCT.Read.qmd`, `XCT.Read.R`, and the `Datafolder` directory are in the repository root. If your data are elsewhere, change `data_path` to that folder.
:::

## Choosing an output

The `output` argument controls the structure returned by `XCT.read()`.

| `output` | What it returns | Typical use |
|----|----|----|
| `"ringwidth"` | Ring-width chronology in `dplR`/RWL-style wide format | Dendrochronological ring-width analyses |
| `"density"` | Annual density parameter in `dplR`/RWL-style wide format | Density chronologies |
| `"ringwidth_density"` | Long table with sample, year, ring width and density | Statistics, plotting and data merging |
| `"density_profile"` | Pixel-level density values within each annual ring | Inspecting within-ring density structure |
| `"MXD"` | MXD chronology read from RingIndicator `*_MXD.rwl` files | Area-of-interest MXD exports |

### Ring width only

```{r, echo = T, results = 'hide'}
#| label: ringwidth-output
RW <- XCT.read(
  path = data_path,
  output = "ringwidth"
)
```

The returned object is convenient for functions in `dplR`: years are row names and individual samples are columns.

### Ring width and density together

For many analyses, `"ringwidth_density"` is the most convenient output because it is already in long format.

```{r, echo = T, results = 'hide'}
#| label: rw-density-output
RW_density <- XCT.read(
  path = data_path,
  output = "ringwidth_density",
  densityType = "fraction",
  area = c(0.75, 1),
  fun = "mean"
)
```

```{r}
head(RW_density) |>
  kable(digits = 4)
```

This example defines density as the **mean density in the final 25% of each ring**.

## Defining the density window

A density statistic is only meaningful after defining **which part of the ring** should be used. `XCT.read()` supports two approaches.

### 1. Fraction of each ring

With `densityType = "fraction"`, the window is defined relative to ring width. The ring runs from `0` at its beginning to `1` at its end.

``` text
Ring start                                                          Ring end
0                                                                   1
|----------------|----------------|----------------|----------------|
0              0.25              0.50             0.75              1
                                                   <--------------- >
                                                   area = c(0.75, 1)
```

For example:

``` r
Density_fraction <- XCT.read(
  path = data_path,
  output = "density",
  densityType = "fraction",
  area = c(0.75, 1),
  fun = "mean"
)
```

Because the window scales with the width of every individual ring, `c(0.75, 1)` always means the last quarter, irrespective of whether a ring is narrow or wide.

Common examples are:

| `area`       | Ring section  |
|--------------|---------------|
| `c(0, 1)`    | complete ring |
| `c(0, 0.5)`  | first half    |
| `c(0.5, 1)`  | second half   |
| `c(0.75, 1)` | final quarter |
| `c(0.9, 1)`  | final 10%     |

### 2. Fixed physical width

With `densityType = "fixed"`, the selected window is a fixed number of micrometres from either the beginning or the end of the ring.

For example, the final 100 µm:

``` r
Density_last_100um <- XCT.read(
  path = data_path,
  output = "density",
  densityType = "fixed",
  area = c("end", 100),
  fun = "mean"
)
```

Or the first 200 µm:

``` r
Density_first_200um <- XCT.read(
  path = data_path,
  output = "density",
  densityType = "fixed",
  area = c("start", 200),
  fun = "mean"
)
```

::: callout-important
## Fraction and fixed windows answer different questions

A **fraction** represents the same relative part of every ring, but its physical width changes with ring width.

A **fixed** window represents the same physical distance in every ring, but it occupies a larger fraction of narrow rings than of wide rings.
:::

## Choosing the density statistic

The `fun` argument determines how the density values inside the selected window are reduced to one value per ring.

| `fun` | Calculation | Example use |
|----|----|----|
| `"mean"` | arithmetic mean | Average earlywood/latewood density |
| `"median"` | median | More robust central density |
| `"min"` | minimum | Minimum density in the selected region |
| `"max"` | maximum | Maximum observed density |
| `"mean_top_x"` | mean of the highest fraction of values | A less pixel-sensitive high-density metric |

### Mean of the highest values

`"mean_top_x"` uses the additional argument `x`. For example, this calculates the mean of the highest 20% of density values found in the final quarter of each ring:

``` r
Density_top20 <- XCT.read(
  path = data_path,
  output = "density",
  densityType = "fraction",
  area = c(0.75, 1),
  fun = "mean_top_x",
  x = 0.20
)
```

This differs from `fun = "max"`: a maximum is determined by one value, whereas `mean_top_x` summarizes a group of high values.

## Removing very narrow rings from density calculations

Very narrow rings may contain too few pixels for a meaningful density statistic. They can be excluded from the **density calculation** using `removeNarrowRings` and `minRingWidth`.

``` r
Data_filtered <- XCT.read(
  path = data_path,
  output = "ringwidth_density",
  densityType = "fraction",
  area = c(0.75, 1),
  fun = "mean",
  removeNarrowRings = TRUE,
  minRingWidth = 0.030
)
```

`minRingWidth` is expressed in **mm**. In this example, density is not calculated for rings narrower than 0.030 mm.

::: callout-note
With `removeNarrowRings = TRUE`, rings below the threshold are excluded from the density calculation. If you also need the complete ring-width chronology, read `output = "ringwidth"` separately and join it to the density result by year/sample as needed.
:::

## Resolution handling

Ring width is stored in pixels in the RingIndicator export and converted to physical width using the reported pixel size.

By default, `XCT.read()` uses the resolution stored in the `*_ringwidth.txt` files.

``` r
Data <- XCT.read(
  path = data_path,
  output = "ringwidth_density",
  overruleResolution = FALSE
)
```

### Automatically handling suspicious resolution values

With the default `autoFixWeirdResolution = TRUE`, the function checks for strongly deviating resolution values and can replace obvious factor-of-ten outliers with the most common resolution found among samples.

``` r
Data <- XCT.read(
  path = data_path,
  output = "ringwidth_density",
  autoFixWeirdResolution = TRUE,
  verbose = TRUE
)
```

Keep `verbose = TRUE` when first loading a dataset so that the resolution summary and possible warnings are visible.

### Manually overriding resolution

If you know that all profiles should use a specific resolution, you can force it:

``` r
Data <- XCT.read(
  path = data_path,
  output = "ringwidth_density",
  overruleResolution = TRUE,
  resolution = 1
)
```

Here, `resolution = 1` means **1 µm per pixel**.

::: callout-warning
Only override the reported resolution when you know the correct voxel/pixel size. A wrong resolution directly changes calculated ring widths and fixed-width density windows.
:::

## Getting the complete density profile

Use `output = "density_profile"` when you want to analyse or visualise the density variation inside annual rings rather than reducing each ring to a single value.

``` r
Profile <- XCT.read(
  path = data_path,
  output = "density_profile"
)
```

The returned long table contains the sample, year, pixel number along the ring, and density.

## Reading MXD files

When maximum latewood density has already been extracted in RingIndicator using areas of interest, use:

``` r
MXD <- XCT.read(
  path = data_path,
  output = "MXD"
)

head(MXD)
```

This output uses the sample-specific `*_MXD.rwl` files. The combined `ALL_MXD.rwl` file is not needed for this step.

## Function arguments at a glance

```{r}
#| label: argument-table
#| echo: false

tribble(
  ~Argument, ~Purpose, ~Typical_value,
  "path", "Folder containing RingIndicator export files", '"Datafolder"',
  "output", "Choose the returned data structure", '"ringwidth_density"',
  "densityType", "Use a relative or physical density window", '"fraction" or "fixed"',
  "area", "Start/end of the density window", 'c(0.75, 1) or c("end", 100)',
  "fun", "Statistic calculated inside the window", '"mean"',
  "x", "Highest fraction used by mean_top_x", "0.20",
  "removeNarrowRings", "Exclude narrow rings from density calculations", "FALSE",
  "minRingWidth", "Minimum ring width when filtering (mm)", "0.030",
  "overruleResolution", "Force one resolution for all samples", "FALSE",
  "resolution", "Forced resolution (µm/pixel)", "1",
  "autoFixWeirdResolution", "Correct obvious factor-of-ten resolution outliers", "TRUE",
  "verbose", "Print loading and resolution information", "TRUE"
) |>
  kable(align = c("l", "l", "l"))
```

## A complete analysis example

The following pattern is a good starting point for an analysis script.

``` r
library(tidyverse)
library(dplR)
source("XCT.Read.R")

Data <- XCT.read(
  path = "Datafolder",
  output = "ringwidth_density",
  densityType = "fraction",
  area = c(0.75, 1),
  fun = "mean",
  removeNarrowRings = TRUE,
  minRingWidth = 0.030,
  verbose = TRUE
)

# Inspect the result
summary(Data)

# Missing density values after filtering
Data |>
  filter(is.na(Density))

# Plot one sample
sample_to_plot <- Data |>
  count(Sample, sort = TRUE) |>
  slice(1) |>
  pull(Sample)

Data |>
  filter(Sample == sample_to_plot) |>
  ggplot(aes(Year, Density)) +
  geom_line() +
  labs(
    title = paste("Density series:", sample_to_plot),
    x = "Year",
    y = "Density"
  ) +
  theme_minimal()
```

## Troubleshooting

### "No eligible ring width files found"

Check that the folder contains files ending in `_ringwidth.txt` and that `path` points to the correct directory.

### Density output cannot be calculated

For the standard density outputs, each sample should have matching corrected density and ring-position files, normally:

``` text
<sample>_density_corr.txt
<sample>_zpos_corr.txt
<sample>_ringwidth.txt
```

### Some samples disappear

Check the messages printed with `verbose = TRUE`. Differences between sample names across file groups or incomplete exports can prevent matching.

### Density values are missing for narrow rings

If `removeNarrowRings = TRUE`, this may be intentional. Compare the ring width with `minRingWidth`.

### Ring widths appear ten times too large or too small

Inspect the resolution summary. Verify the `pixelsize` values in the RingIndicator export and the settings of `autoFixWeirdResolution`, `overruleResolution`, and `resolution`.

## Citation and further information

`XCT.read()` is part of the UGent-Woodlab X-ray micro-CT tree-ring densitometry workflow.

- Project website: <https://dendrochronomics.ugent.be/>
- XCT.read GitHub repository: <https://github.com/UGent-Woodlab/XCT.read-R-function>
- XCT.read Zenodo DOI: <https://doi.org/10.5281/zenodo.14654939>
- Pipeline paper: <https://doi.org/10.1016/j.dendro.2025.126343>

Please follow the citation guidance in the repository README and on the Dendrochronomics website when using the software in published work.
